Appendix

### 2025-10-17 manuscript_rmarkdown/manuscript_appendix.Rmd
Add information on subsets of ranking of Yeo-Johnson based methods.

>>>>

## Ranking Yeo-Johnson transformations (pt. II)

The main manuscript shows the overall ranking of different Yeo-Johnson-based power transformation methods across three
sets of problems.
To assess the effect of limited sequence lengths, we perform ranking on subsets of the experiment.
The results for ranking the subset of sequences with 30 samples or fewer ($n = 16001$) 
is shown in Table \ref{tab:comparison_methods_simulations_yeo_johnson_10_30}.
In the clean dataset without outliers, where all sequences were sampled from a normal distribution and 
then underwent an inverse Yeo-Johnson transformation,
conventional and location- and scale-invariant are ranked similarly, with no clear advantage
to any method. For the remaining datasets, the robust location- and scale-invariant
method ranked best. It slightly outperformed (robustly scaled) conventional Yeo-Johnson transformations
and (z-score normalised) robust transformations [@Raymaekers2024-zf], in datasets without and with outliers, respectively.

In the subset for larger sequence lengths, with at least 1000 samples per sequence ($n = 33037$),
the cost of making the location- and scale-invariant transformation robust against outliers
becomes more apparent (Table \ref{tab:comparison_methods_simulations_yeo_johnson_1000_10000}).
In the clean dataset without outliers, down-weighting of the tails of the distribution by
the robust location- and scale-invariant method leads to larger residuals compared 
to other methods (Table \ref{tab:comparison_methods_simulations_yeo_johnson_1000_10000_avg_residual}.

# Yeo-Johnson on clean and dirty data.
clean_data <- .assess_simulated_data(manuscript_dir = manuscript_dir, data_type = "clean")
dirty_data <- .assess_simulated_data(manuscript_dir = manuscript_dir, data_type = "dirty")
shifted_data <- .assess_simulated_data(manuscript_dir = manuscript_dir, data_type = "dirty_shifted")

# 30 samples or fewer
data <- data[n_samples <= 30L, ]

# Add rank based on residual error for central normality.
data[outlier == FALSE, "method_rank" := rank(residuals), by = c("experiment_id", "outlier", "dataset")]
data[outlier == TRUE, "method_rank" := rank(residuals_central), by = c("experiment_id", "outlier", "dataset")]
data[outlier == FALSE, "norm_residuals" := residuals / n_samples]
data[outlier == TRUE, "norm_residuals" := residuals_central / n_samples]

table_data <- data[
  ,
  list(
    "avg." = round(mean(method_rank), digits = 2L),
    "10%" = quantile(method_rank, 0.10),
    "50%" = quantile(method_rank, 0.50),
    "90%" = quantile(method_rank, 0.90),
    "avg. norm. res." = round(mean(norm_residuals), digits = 3)
  ),
  by = c("transformation", "outlier", "dataset")
]

# Table with average rank.
data.table::dcast(transformation ~ dataset + outlier, data = table_data, value.var = "avg.")

# Table with average normalised residuals.
data.table::dcast(transformation ~ dataset + outlier, data = table_data, value.var = "avg. norm. res.")

\begin{table}
\begin{center}
\caption{
Comparison of average rank between Yeo-Johnson transformation methods based on either residual error (without outliers) or residual error of the central portion
(with outliers; $\kappa = 0.80$) over 3 datasets with 16001 sequences each. Each sequence contained 30 samples or fewer. 
The clean dataset consists of sequences derived through inverse Yeo-Johnson transformation
of data sampled from a standard normal distribution. The dirty dataset contains sequences sampled from asymmetric generalised normal distributions, centred at 0.
The shifted dataset also contains sequences sampled from asymmetric generalised normal distributions, but centred at 100, and scaled by 0.001.
Several transformation methods include normalisation before transformation, indicated by z-score normalisation (norm.) or robust scaling.
A rank of 1 is the best and a rank of 9 the worst. For each dataset, the best ranking transformation is marked in bold.
}
\label{tab:comparison_methods_simulations_yeo_johnson_10_30}
\begin{tabular}{l | l r r r r r r}

none                                  & &         7.75  &         6.81  &         7.27  &         6.40  &         7.45  &         6.60 \\
conventional                          & &         4.19  &         5.47  &         4.48  &         5.78  &         6.55  &         5.94 \\
conventional (z-score norm.)          & &         4.36  &         5.12  &         5.01  &         4.99  &         4.35  &         4.70 \\
conventional (robust scaling)         & & \textbf{4.05} &         5.63  &         4.38  &         5.49  &         3.76  &         5.20 \\
Raymaekers-Rousseeuw                  & &         5.36  &         4.19  &         5.14  &         4.62  &         6.32  &         5.83 \\
Raymaekers-Rousseeuw (z-score norm.)  & &         5.56  &         4.08  &         5.67  &         4.05  &         5.03  &         3.81 \\
Raymaekers-Rousseeuw (robust scaling) & &         5.33  &         4.14  &         5.15  &         4.15  &         4.55  &         3.94 \\
invariant                             & &         4.11  &         5.92  &         4.12  &         5.67  &         3.62  &         5.37 \\
robust invariant                      & &         4.28  & \textbf{3.65} & \textbf{3.79} & \textbf{3.85} & \textbf{3.37} & \textbf{3.61} \\

\begin{table}
\begin{center}
\caption{
Comparison of average normalised residual error between Yeo-Johnson transformation methods based on either the residual error (without outliers) or the residual error of the central portion
(with outliers; $\kappa = 0.80$) over 3 datasets with 16001 sequences each. Residual errors were normalised by sequence length to reduce the effect of sequence length on the residual error.
Each sequence contained 30 samples or fewer. The clean dataset consists of sequences derived through inverse Yeo-Johnson transformation
of data sampled from a standard normal distribution. The dirty dataset contains sequences sampled from asymmetric generalised normal distributions, centred at 0.
The shifted dataset also contains sequences sampled from asymmetric generalised normal distributions, but centred at 100, and scaled by 0.001.
Several transformation methods include normalisation before transformation, indicated by z-score normalisation (norm.) or robust scaling.
A lower normalised residual error is better. Discrepancies with the rank-based method may occur, as even with normalisation residual errors 
are not fully comparable between experiments, whereas the ranks achieved within each experiment are.
}
\label{tab:comparison_methods_simulations_yeo_johnson_10_30_avg_residual}
\begin{tabular}{l | l r r r r r r}

none                                  & & 0.246 & 0.139 & 0.186 & 0.125 & 0.187 & 0.125 \\
conventional                          & & 0.139 & 0.114 & 0.142 & 0.106 & 0.187 & 0.125 \\
conventional (z-score norm.)          & & 0.139 & 0.113 & 0.142 & 0.104 & 0.142 & 0.104 \\
conventional (robust scaling)         & & 0.139 & 0.114 & 0.141 & 0.104 & 0.141 & 0.104 \\
Raymaekers-Rousseeuw                  & & 0.211 & 0.121 & 0.183 & 1.212 & 0.187 & 0.125 \\
Raymaekers-Rousseeuw (z-score norm.)  & & 0.295 & 0.129 & 0.175 & 0.125 & 0.175 & 0.125 \\
Raymaekers-Rousseeuw (robust scaling) & & 0.209 & 0.124 & 0.172 & 0.116 & 0.172 & 0.116 \\
invariant                             & & 0.141 & 0.115 & 0.140 & 0.106 & 0.140 & 0.106 \\
robust invariant                      & & 0.141 & 0.100 & 0.140 & 0.095 & 0.140 & 0.090 \\

# Yeo-Johnson on clean and dirty data.
clean_data <- .assess_simulated_data(manuscript_dir = manuscript_dir, data_type = "clean")
dirty_data <- .assess_simulated_data(manuscript_dir = manuscript_dir, data_type = "dirty")
shifted_data <- .assess_simulated_data(manuscript_dir = manuscript_dir, data_type = "dirty_shifted")

# 30 samples or fewer
data <- data[n_samples >= 1000L, ]

# Add rank based on residual error for central normality.
data[outlier == FALSE, "method_rank" := rank(residuals), by = c("experiment_id", "outlier", "dataset")]
data[outlier == TRUE, "method_rank" := rank(residuals_central), by = c("experiment_id", "outlier", "dataset")]
data[outlier == FALSE, "norm_residuals" := residuals / n_samples]
data[outlier == TRUE, "norm_residuals" := residuals_central / n_samples]

table_data <- data[
  ,
  list(
    "avg." = round(mean(method_rank), digits = 2L),
    "10%" = quantile(method_rank, 0.10),
    "50%" = quantile(method_rank, 0.50),
    "90%" = quantile(method_rank, 0.90),
    "avg. norm. res." = round(mean(norm_residuals), digits = 3)
  ),
  by = c("transformation", "outlier", "dataset")
]

# Table with average rank.
data.table::dcast(transformation ~ dataset + outlier, data = table_data, value.var = "avg.")

# Table with average normalised residuals.
data.table::dcast(transformation ~ dataset + outlier, data = table_data, value.var = "avg. norm. res.")

\begin{table}
\begin{center}
\caption{
Comparison of average rank between Yeo-Johnson transformation methods based on either residual error (without outliers) or residual error of the central portion
(with outliers; $\kappa = 0.80$) over 3 datasets with 33037 sequences each. Each sequence contained between 1000 and 10000 samples.
The clean dataset consists of sequences derived through inverse Yeo-Johnson transformation
of data sampled from a standard normal distribution. The dirty dataset contains sequences sampled from asymmetric generalised normal distributions, centred at 0.
The shifted dataset also contains sequences sampled from asymmetric generalised normal distributions, but centred at 100, and scaled by 0.001.
Several transformation methods include normalisation before transformation, indicated by z-score normalisation (norm.) or robust scaling.
A rank of 1 is the best and a rank of 9 the worst. For each dataset, the best ranking transformation is marked in bold.
}
\label{tab:comparison_methods_simulations_yeo_johnson_1000_10000}
\begin{tabular}{l | l r r r r r r}

none                                  & &         8.85  &         8.14  &         7.35  &         6.29  &         7.41  &         6.87 \\
conventional                          & &         3.50  &         5.65  &         3.52  &         7.78  &         6.42  &         6.05 \\
conventional (z-score norm.)          & &         5.24  &         5.94  &         7.34  &         4.98  &         6.26  &         4.79 \\
conventional (robust scaling)         & &         3.53  &         6.98  &         5.97  &         5.79  &         4.87  &         5.65 \\
Raymaekers-Rousseeuw                  & &         4.15  &         2.81  &         3.37  &         3.72  &         6.19  &         5.79 \\
Raymaekers-Rousseeuw (z-score norm.)  & &         5.34  & \textbf{1.86} &         6.57  &         3.39  &         5.49  &         3.30 \\
Raymaekers-Rousseeuw (robust scaling) & &         4.20  &         2.33  &         5.21  &         3.36  &         4.12  &         3.23 \\
invariant                             & & \textbf{2.36} &         7.59  & \textbf{2.66} &         6.82  & \textbf{1.98} &         6.81 \\
robust invariant                      & &         7.83  &         3.70  &         3.01  & \textbf{2.88} &         2.26  & \textbf{2.51} \\

\begin{table}
\begin{center}
\caption{
Comparison of average normalised residual error between Yeo-Johnson transformation methods based on either the residual error (without outliers) or the residual error of the central portion
(with outliers; $\kappa = 0.80$) over 3 datasets with 16001 sequences each. Residual errors were normalised by sequence length to reduce the effect of sequence length on the residual error.
Each sequence contained between 1000 and 10000 samples. The clean dataset consists of sequences derived through inverse Yeo-Johnson transformation
of data sampled from a standard normal distribution. The dirty dataset contains sequences sampled from asymmetric generalised normal distributions, centred at 0.
The shifted dataset also contains sequences sampled from asymmetric generalised normal distributions, but centred at 100, and scaled by 0.001.
Several transformation methods include normalisation before transformation, indicated by z-score normalisation (norm.) or robust scaling.
A lower normalised residual error is better. Discrepancies with the rank-based method may occur, as even with normalisation residual errors 
are not fully comparable between experiments, whereas the ranks achieved within each experiment are.
}
\label{tab:comparison_methods_simulations_yeo_johnson_1000_10000_avg_residual}
\begin{tabular}{l | l r r r r r r}

none                                  & & 0.186 & 0.073 & 0.106 & 0.043 & 0.106 & 0.043 \\
conventional                          & & 0.014 & 0.060 & 0.060 & 0.046 & 0.106 & 0.043 \\
conventional (z-score norm.)          & & 0.021 & 0.061 & 0.070 & 0.040 & 0.070 & 0.040 \\
conventional (robust scaling)         & & 0.014 & 0.061 & 0.069 & 0.040 & 0.069 & 0.040 \\
Raymaekers-Rousseeuw                  & & 0.015 & 0.017 & 0.060 & 0.024 & 0.106 & 0.043 \\
Raymaekers-Rousseeuw (z-score norm.)  & & 0.021 & 0.016 & 0.070 & 0.024 & 0.070 & 0.024 \\
Raymaekers-Rousseeuw (robust scaling) & & 0.015 & 0.016 & 0.068 & 0.024 & 0.068 & 0.024 \\
invariant                             & & 0.013 & 0.067 & 0.059 & 0.044 & 0.059 & 0.044 \\
robust invariant                      & & 0.034 & 0.034 & 0.061 & 0.026 & 0.061 & 0.026 \\
